How to select predictive models for decision-making or causal inference?

### 2023-12-17 how_to_select_causal_models.tex
iter

Progress in machine learning promises to integrate new data for
precision medicine \cite{beam2018big,rajkomar2019machine}:
health predictions building on medical images
\cite{khojaste2022deep,zhang2019radiological,yala2019deep,shen2019deep,nassif2022breast},
patient records
\cite{mooney2018bigdata,desai2020comparison,simon2018predicting} or clinical
notes \cite{horng2017creating,wang2020prediction,spasic2020clinical}. Data may
be difficult to control and model, but the accuracy of the prediction can be
verified on left-out data
\cite{altman2009prognosis,poldrack2020establishment,varoquaux2022evaluating}.
Given a model predicting a health outcome, precision medicine would like it to
guide decisions: will an individual benefit from an intervention such as surgery
\cite{fontana2019can}? Contrasting predictions with and without the treatment
gives an answer, but statistical validity requires causal inference
\cite{snowden_implementation_2011,sperrin2019explicit,blakely2020reflection}.

>>>>

Prediction models have long been used in medicine, as with risk
score or prognostic models \cite{moons2009prognosis,steyerberg2019clinical}.
While these have historically been simple models on simple data, this is
changing with progress in machine learning and richer sources of medical data
\cite{beam2018big,rajkomar2019machine}. Health predictions can now integrate medical images
\cite{khojaste2022deep,zhang2019radiological,yala2019deep,shen2019deep,nassif2022breast},
patient records
\cite{mooney2018bigdata,desai2020comparison,simon2018predicting} or clinical
notes \cite{horng2017creating,wang2020prediction,spasic2020clinical}.
Complex data is difficult to control and model, but these models are
validated by verifying the accuracy of the prediction on left-out data
\cite{altman2009prognosis,poldrack2020establishment,varoquaux2022evaluating}.
Given a model predicting a health outcome, precision medicine would like it to
guide decisions: will an individual benefit from an intervention such as surgery
\cite{fontana2019can}? Contrasting predictions with and without the treatment
gives an answer, but statistical validity requires causal inference
\cite{snowden_implementation_2011,sperrin2019explicit,blakely2020reflection}.
